### 2020-08-12 latex/vnhcs.tex
Finished all VNHC proofs
Fixed the errors that were stopping compilation and finished the proofs
in the VNHC chapter. I think I made an error in finding the closed loop
dynamics for general VNHCs, which needs to be discussed with Manfredi.

% TODO: Replace [p D_i p] here with our new notation
The resulting Hamiltonian system reduces to (\ref{eqn:hamiltonian}).
\begin{align}\label{eqn:hamiltonian}
    \mathcal{H}(q,p) &= \frac{1}{2} p\tpose D^{-1}(q) p + P(q) \\
                     &\begin{cases}
        \dot{q} = D\inv(q)p \\
        \dot{p} = -\frac{1}{2} (\Id{n} \otimes p\tpose) \nabla_q D\inv(q) p
        - \nabla_q P(q) + B(q) \tau \\
    \end{cases} 
\end{align}

@@

\section{Simply Actuated Hamiltonian Systems}
% TODO: Describe the change of coordinates to get into q_u/q_a mode and show
% that the actuator directly affects pa but not pu. Use M for simply actuated
% inertia, D for normal coordinates
Given a Hamiltonian mechanical system (\ref{eqn:hamiltonian}), it is not obvious
how the input forces \(\tau\) affect the conjugate of momenta \(p_i\). 
This is because \(\tau\) is transformed by the input matrix \(B(q)\), which may
be quite complicated.

@@

% TODO: Replace the [P M_i P] with new notation
\begin{thm}
    Under Assumptions \ref{assm:B-const},\ref{assm:B-perp}, and
    \ref{assm:B-orthogonal}, the Hamiltonian system (\ref{eqn:hamiltonian}) is
    simply actuated with simply actuated coordinates 
    \(\{Q = \mathbf{B}q, P = \mathbf{B}p\}\). The resulting dynamics are 
    given by (\ref{eqn:simple-hamiltonian}),
    \begin{align}\label{eqn:simple-hamiltonian}
        \hat{H}(Q,P) &= 
        \frac{1}{2} P\tpose \Minv(Q) P + V(Q) \\
       &\begin{cases}
            \dot{Q} = \Minv(Q)P \\
            \dot{P} = -\frac{1}{2} (\Id{n} \otimes P\tpose) \nabla_Q \Minv(Q) P
                - \nabla_Q V(Q) + \simpleB \tau
        \end{cases} \nonumber
    \end{align}
    where
    \begin{align*}
        \Minv(Q) &:= \mathbf{B}D^{-1}(\mathbf{B}\tpose Q)\mathbf{B}\tpose \\
        V(Q) &:= P(\mathbf{B}\tpose Q) \\
    \end{align*}
\end{thm}
\begin{proof}
    The Poisson bracket between the functions \(f(q,p)\) and \(g(q,p)\) is defined by
    \cite{landau_mechanics} as follows:
    \[
        [f,g] := \sum \limits_{i=1}^n \pdiff{f}{p_i}\pdiff{g}{q_i} - 
                \pdiff{f}{q_i}\pdiff{g}{p_i}
    \]
    For any constant matrix \(A\), the transformation
    \(\{Q = Aq, P = Ap\}\) satisfies
    \(\pdiff{Q_i}{p_m} = \pdiff{P_i}{q_m} = 0\) for all 
    \(i,m \in \n\).
    Hence, for our new coordinates \(\{Q = \mathbf{B}q, P = \mathbf{B}p\}\),
    \begin{align*}
        [Q_i,Q_j] &:= \sum\limits_{m = 1}^n \pdiff{Q_i}{p_m}\pdiff{Q_j}{q_m} - 
        \pdiff{Q_i}{q_m}\pdiff{Q_j}{p_m} = 0 \\
        [P_i,P_j] &:= \sum\limits_{m=1}^n \pdiff{P_i}{p_m}\pdiff{P_j}{q_m} -
        \pdiff{P_i}{q_m}\pdiff{P_j}{p_m} = 0
    \end{align*}
    Since the matrix \(\mathbf{B}\) is orthogonal,
    \((\mathbf{B}_i)\tpose \mathbf{B}\tpose_j = (\mathbf{B}_i)\tpose (\mathbf{B}^{-1})_j = \delta_{i,j}\). Using this
    fact we see that the Poisson brackets between \(P\) and \(Q\) are given by:
    \begin{align*}
        [P_i,Q_j] &= \sum\limits_{m=1}^n\pdiff{P_i}{p_m}\pdiff{Q_j}{q_m}
        - \pdiff{P_i}{q_m}\pdiff{Q_j}{p_m} \\
                  &= \sum\limits_{m=1}^n \mathbf{B}_{i,m}\mathbf{B}_{j,m} - 0 \\
                  &= \sum\limits_{m=1}^n \mathbf{B}_{i,m}\mathbf{B}\tpose_{m,j} \\
                  &= (\mathbf{B}_i)\tpose \mathbf{B}\tpose_j \\
                  &= \delta_{i,j}
    \end{align*}

@@

\section{Virtual Nonholonomic Constraints}
% TODO: Motivation for VNHCs, they are an extension of VHCs, etc.
Let us imagine a child on a swing, who wants to go as high as possible. The
child will push off the ground and start swinging with small oscillations,
then extend and retract their feet appropriately to gain energy. 
If a roboticist were designing a machine to do this task, they might design a
control mechanism which makes the legs track a trajectory over time. For
example, they might tell the robot to extend and retract its legs every two
seconds. In ideal situtations, this technique would work perfectly because the
leg motion is synchronized with the swing to gain energy as fast as possible.

@@

% TODO: Perform the full development of VNHCs, including its stabilizing
% controller.
\begin{defn}
    A \textit{virtual nonholonomic constraint} (VNHC) \textit{of order \(k\)} is a
    relation \(h(q,p) = 0\) where \(h : \mathcal{Q}\times\R^n \rightarrow \R^k\) is
    smooth, \(\rank{\left[ dh_q,\, dh_p \right]} = k\) for all 
    \((q,p) \in h\inv(0)\), and there exists a feedback controller \(\tau(q,p)\)
    stabilizing the set
    \[
        \Gamma = \left\{(q,p) \mid h(q,p) = 0, dh_q \dot{q} + dh_p \dot{p} = 0\right\}
    \]
    which is called the \textit{constraint manifold}.
\end{defn}

@@

    % TODO: complete this proof with our notation

@@

@@

    For the next segment, let \(e_j \in \R^n\) be the vector of all zeros,

@@

        = \sum\limits_{j=1}^n \pdiff{\Minv}{q_j}(q)\left(e_j\tpose\Minv(q)p\right)

@@

@@

    Defining \(C_2 : \mathcal{Q} \times \mathcal{P} \rightarrow \R^{n(n-k)\times n}\)
    allows us to write the entire derivative in matrix form:

@@

    Finally, we compute the derivative of \(\nabla_{q_u}V(q)\)

@@

        \end{bmatrix = \partial_{q_u} \partial_q V(q) \dot{q}
    and so we get that
        \partial_{q_u} \partial_q V(q) \Minv(q)p

@@

     \frac{1}{2}dh_{p_u}(\Id{n-k}\otimes p\tpose)C_2(q,p)p - 

@@

     \left(dh_q \Minv(q) - dh_{p_u}(\Id{n-k} \otimes

@@

    The VNHC \(h\) is regular iff \(e\) is of relative degree
    \(\{2,\ldots,2\}\), which is true iff \(\tau\) appears in each element of
    \(e\) on the constraint manifold. 
    Letting \(\ddot{e} = E(q,p) + H(q,p)\tau\), this is equivalent to requiring
    that the matrix \(H\) be full rank.

@@

%TODO: Special case: when dM/dqu = 0, show that we have a nice form and that we
%can solve for p_a and the closed-loop dynamics (qu,pu)_dot

Now that we have found a controller to enforce a regular VNHC of order \(k\), we
would like to solve for the closed-loop dynamics. Naturally, these dynamics
should be specified through \((q_u, p_u)\) since \(q_a\) is a function of these
as specified by \(h(q,p_u) = 0\).
Unfortunately, \(\dot{q}_u\) depends on \(p_a\), and for general systems one
cannot solve explicitly for \(p_a\) in terms of \((q_u,p_u)\). This is because
the \(\dot{p}\) dynamics contains the coupling term 
\((\Id{n} \otimes p\tpose)\nabla_{q_u}M(q)p\).

@@

\begin{thm}\label{thm:zero-dynamics}
    Let \(\mathcal{H}\) be a ----- mechanical system satisfying Assumption
    \ref{assm:H-is-simply-actuated}. Let
    \(h(q,p_u) = 0\) be a regular VNHC of order \(k\) with constraint manifold
    \(\Gamma\).
    Then, on \(\Gamma\), the closed-loop dynamics are given by
    \begin{equation}\label{eqn:qpu-dynamics}
        \left.\begin{aligned}
            \dot{q}_u &= M(q)p \\
            \dot{p}_u &= -\nabla_{q_u}V(q) \\
            \end{aligned}\right|_{\begin{array}{c}
                h(q,p_u) = 0 \\ 
                pa = g(q_u,p_u) \\
            \end{array}}
    \end{equation}
    where
    \begin{equation}\label{eqn:g-qpu}
    \begin{aligned}
        g(q,p_u) &:= \\
           &\left.\left(dh_q \Minv(q) \simpleB \right)\inv 
        \left(dh_{p_u} \nabla_{q_u}V(q) - dh_q \Minv(q)
        \begin{bmatrix}
            I_{n-k} \\
            \Zmat{k \times (n-k)} \\
        \end{bmatrix} p_u\right)\right|_{h(q,p_u) = 0}
    \end{aligned}
    \end{equation}
\end{thm}
\begin{proof}
%TODO: Complete the proof for this theorem
\end{proof}

The most fascinating part of Theorem \ref{thm:zero-dynamics} is that it shows
that the dynamics on \(\Gamma\) are entirely described by the \(2(n-k)\)
unactuated coordinates, regardless of the number of degrees of freedom of the
system. This means that \(\Gamma\) is a \(2(n-k)\)-dimensional surface from
which the system cannot escape. This becomes especially important when analyzing
systems of degree of underactuation one, since it means the constraint is always
a curve which can be plotted on the plane with axes \(q_u\) and \(p_u\)
regardless of the original dimension \(n\) of the system.

In future sections of this thesis, we will apply a very specific type of
constraint. The following corollay applies Theorem \ref{thm:zero-dynamics} to
these constraints.

\begin{cor}\label{cor:2d-zero-dynamics}
    Suppose \(\mathcal{H}\) is a ---- mechanical system satisfying Assumption
    \ref{assm:H-is-simply-actuated} which has degree of underactuation one.
    Let \(h(q,p_u) = q_a - f(q_u,p_u)\) be a regular VNHC of order \((n-1)\),
    where \(f\) is a suitably-defined smooth function.
    Defining \(e_1 := (1,0,\ldots,0) \in \R^n\), the actuated momentum 
    is given by
    \begin{equation}\label{eqn:g-qupu}
        p_a = -\left.\left(dh_q \Minv(q)
        \begin{bmatrix}
            \Zmat{1\times (n-1)}\\
            I_{(n-1)}
        \end{bmatrix}\right)\inv 
        \left(\partial_{p_u}f \partial_{q_u}V + dh_q \Minv(q)e_1 p_u\right) 
            \right|_{q_a =f(q_u,p_u)}
    \end{equation}
    Since \(q_u \in \Rt{T}\) for some \(T > 0\), \((q_u(t),p_u(t))\) traces out
    a curve on the so-called \((q,p)\)-plane \(\Rt{T} \times \R\).
\end{cor}

>>>>

The resulting Hamiltonian system reduces to (\ref{eqn:hamiltonian}).
\begin{align}\label{eqn:hamiltonian}
    \mathcal{H}(q,p) &= \frac{1}{2} p\tpose D^{-1}(q) p + P(q) \\
                     &\begin{cases}
        \dot{q} = D\inv(q)p \\
        \dot{p} = -\frac{1}{2} (\Id{n} \otimes p\tpose) \nabla_q D\inv(q) p
        - \nabla_q P(q) + B(q) \tau \\
    \end{cases} 
\end{align}

@@

\section{Simply Actuated Hamiltonian Systems}
% Describe the change of coordinates to get into q_u/q_a mode and show
% that the actuator directly affects pa but not pu. Use M for simply actuated
% inertia, D for normal coordinates
Given a Hamiltonian mechanical system (\ref{eqn:hamiltonian}), it is not obvious
how the input forces \(\tau\) affect the conjugate of momenta \(p_i\). 
This is because \(\tau\) is transformed by the input matrix \(B(q)\), which may
be quite complicated.

@@

\begin{thm}
    Under Assumptions \ref{assm:B-const},\ref{assm:B-perp}, and
    \ref{assm:B-orthogonal}, the Hamiltonian system (\ref{eqn:hamiltonian}) is
    simply actuated with simply actuated coordinates 
    \(\{Q = \mathbf{B}q, P = \mathbf{B}p\}\). The resulting dynamics are 
    given by (\ref{eqn:simple-hamiltonian}),
    \begin{align}\label{eqn:simple-hamiltonian}
        \hat{H}(Q,P) &= 
        \frac{1}{2} P\tpose \Minv(Q) P + V(Q) \\
       &\begin{cases}
            \dot{Q} = \Minv(Q)P \\
            \dot{P} = -\frac{1}{2} (\Id{n} \otimes P\tpose) \nabla_Q \Minv(Q) P
                - \nabla_Q V(Q) + \simpleB \tau
        \end{cases} \nonumber
    \end{align}
    where
    \begin{align*}
        \Minv(Q) &:= \mathbf{B}D^{-1}(\mathbf{B}\tpose Q)\mathbf{B}\tpose \\
        V(Q) &:= P(\mathbf{B}\tpose Q) \\
    \end{align*}
\end{thm}
\begin{proof}
    The Poisson bracket between the functions \(f(q,p)\) and \(g(q,p)\) is defined by
    \cite{landau_mechanics} as follows:
    \[
        [f,g] := \sum \limits_{i=1}^n \pdiff{f}{p_i}\pdiff{g}{q_i} - 
                \pdiff{f}{q_i}\pdiff{g}{p_i}
    \]
    For any constant matrix \(A\), the transformation
    \(\{Q = Aq, P = Ap\}\) satisfies
    \(\pdiff{Q_i}{p_m} = \pdiff{P_i}{q_m} = 0\) for all 
    \(i,m \in \n\).
    Hence, for our new coordinates \(\{Q = \mathbf{B}q, P = \mathbf{B}p\}\),
    \begin{align*}
        [Q_i,Q_j] &:= \sum\limits_{m = 1}^n \pdiff{Q_i}{p_m}\pdiff{Q_j}{q_m} - 
        \pdiff{Q_i}{q_m}\pdiff{Q_j}{p_m} = 0 \\
        [P_i,P_j] &:= \sum\limits_{m=1}^n \pdiff{P_i}{p_m}\pdiff{P_j}{q_m} -
        \pdiff{P_i}{q_m}\pdiff{P_j}{p_m} = 0
    \end{align*}
    Since the matrix \(\mathbf{B}\) is orthogonal,
    \((\mathbf{B}_i)\tpose \mathbf{B}\tpose_j = (\mathbf{B}_i)\tpose (\mathbf{B}^{-1})_j = \delta_{i,j}\). Using this
    fact we see that the Poisson brackets between \(P\) and \(Q\) are given by:
    \begin{align*}
        [P_i,Q_j] &= \sum\limits_{m=1}^n\pdiff{P_i}{p_m}\pdiff{Q_j}{q_m}
        - \pdiff{P_i}{q_m}\pdiff{Q_j}{p_m} \\
                  &= \sum\limits_{m=1}^n \mathbf{B}_{i,m}\mathbf{B}_{j,m} - 0 \\
                  &= \sum\limits_{m=1}^n \mathbf{B}_{i,m}\mathbf{B}\tpose_{m,j} \\
                  &= (\mathbf{B}_i)\tpose \mathbf{B}\tpose_j \\
                  &= \delta_{i,j}
    \end{align*}

@@

\section{Virtual Nonholonomic Constraints}
% Motivation for VNHCs, they are an extension of VHCs, etc.
Let us imagine a child on a swing, who wants to go as high as possible. The
child will push off the ground and start swinging with small oscillations,
then extend and retract their feet appropriately to gain energy. 
If a roboticist were designing a machine to do this task, they might design a
control mechanism which makes the legs track a trajectory over time. For
example, they might tell the robot to extend and retract its legs every two
seconds. In ideal situtations, this technique would work perfectly because the
leg motion is synchronized with the swing to gain energy as fast as possible.

@@

% Perform the full development of VNHCs, including its stabilizing
% controller.
\begin{defn}
    A \textit{virtual nonholonomic constraint} (VNHC) \textit{of order \(k\)} is a
    relation \(h(q,p) = 0\) where \(h : \mathcal{Q}\times\R^n \rightarrow \R^k\) is
    smooth, \(\rank{\left[ dh_q,\, dh_p \right]} = k\) for all 
    \((q,p) \in h\inv(0)\), and there exists a feedback controller \(\tau(q,p)\)
    stabilizing the set
    \[
        \Gamma = \left\{(q,p) \mid h(q,p) = 0, dh_q \dot{q} + dh_p \dot{p} = 0\right\}
    \]
    which is called the \textit{constraint manifold}.
\end{defn}

@@

@@

% d/dt(dh_q)

@@

% C_1(q,p)
    For the next segment, let \(1_j \in \R^n\) be the vector of all zeros,

@@

        = \sum\limits_{j=1}^n \pdiff{\Minv}{q_j}(q)\left(1_j\tpose\Minv(q)p\right)

@@

% d/dt(dh_pu)

@@

    Define \(C_2 : \mathcal{Q} \times \mathcal{P} \rightarrow \R^{n(n-k)\times n}\) 
    by
        C_2(q,p) := \begin{bmatrix}
            \sum_{j=1}^n \ppdiff{\Minv}{q_{u_1}}{q_j} (1_j\tpose \Minv(q)p) \\
            \sum_{j=1}^n \ppdiff{\Minv}{q_{u_{n-k}}}{q_j} (1_j\tpose \Minv(q)p) \\
    This allows us to write the entire derivative in matrix form:

@@

% d/dt(nabla_qu V)
    Finally, we compute the derivative of \(\nabla_{q_u}V(q)\):

@@

        \end{bmatrix} = \partial_{q_u} \partial_q V(q) \dot{q} 
        = \partial_{q_u} \partial_q V(q) \Minv(q)p
% Full ddot(e)

@@

     & \frac{1}{2}dh_{p_u}(\Id{n-k}\otimes p\tpose)C_2(q,p)p - 

@@

     & \left(dh_q \Minv(q) - dh_{p_u}(\Id{n-k} \otimes

@@

% End of Proof
    For shorthand, we'll write \(\ddot{e} = E(q,p) + H(q,p)\tau\) where \(E\)
    and \(H\) are defined appropriately.
    From the definition of regularity, the VNHC \(h\) is regular 
    iff \(e\) is of relative degree \(\{2,\ldots,2\}\), which is true 
    iff \(\tau\) appears in each row of \(\ddot{e}\) on \(\Gamma\). 
    This is equivalent to requiring that the matrix \(H\) be full rank
    everywhere on the constraint manifold, proving the theorem.

@@

% Special case: when dM/dqu = 0, show that we have a nice form and that we
%can solve for p_a and the closed-loop dynamics (qu,pu)_dot

Now that we have found a controller to enforce a regular VNHC of order \(k\), we
would like to solve for the closed-loop dynamics. Intuitively, these dynamics
should be parameterized by \((q_u, p_u)\) since \(q_a\) is a function of these
as specified by \(h(q,p_u) = 0\).
Unfortunately, \(\dot{q}_u\) depends on \(p_a\), and for general systems one
cannot solve explicitly for \(p_a\) in terms of \((q_u,p_u)\). This is because
the \(\dot{p}\) dynamics contains the coupling term 
\((\Id{n} \otimes p\tpose)\nabla_{q_u}M(q)p\).

@@

\begin{thm}\label{thm:zero-dynamics}
    Let \(\mathcal{H}\) be a ----- mechanical system satisfying Assumption
    \ref{assm:H-is-simply-actuated}. Let
    \(h(q,p_u) = 0\) be a regular VNHC of order \(k\) 
    with constraint manifold \(\Gamma\). Suppose that on \(\Gamma\) one can
    solve linearly for \(q_a\) as a function of \(q_u,p_u\).
    Then the closed-loop dynamics are given by
    \begin{equation}\label{eqn:qpu-dynamics}
        \left.\begin{aligned}
            \dot{q}_u &= M(q)p \\
            \dot{p}_u &= -\nabla_{q_u}V(q) \\
            \end{aligned}\right|_{\begin{array}{c}
                h(q,p_u) = 0 \\ 
                pa = g(q_u,p_u) \\
            \end{array}}
    \end{equation}
    where
    \begin{equation}\label{eqn:g-qpu}
    \begin{aligned}
        g(q,p_u) &:= \\
           &\left.\left(dh_q \Minv(q) \simpleB \right)\inv 
        \left(dh_{p_u} \nabla_{q_u}V(q) - dh_q \Minv(q)
        \begin{bmatrix}
            I_{n-k} \\
            \Zmat{k \times (n-k)} \\
        \end{bmatrix} p_u\right)\right|_{h(q,p_u) = 0}
    \end{aligned}
    \end{equation}
\end{thm}
\begin{proof}
    Setting \(e = h(q,p_u)\) and using the fact that 
    \(\nabla_{q_u}\Minv(q) = 0\), we find that
    \[
        \dot{e} = dh_q\Minv(q)p - dh_{p_u}\nabla_{q_u}V(q)
    \]
    Observe that
    \begin{align*}
        dh_q\Minv(q)p &= dh_q\Minv(q) \begin{bmatrix}p_u \\ p_a \end{bmatrix} \\
          &= dh_q\Minv(q) \begin{bmatrix}
              \Id{n-k} & \Zmat{(n-k)\times k} \\
              \Zmat{k \times(n-k)} & \Id{k} 
              \end{bmatrix} \begin{bmatrix} p_u \\ p_a \end{bmatrix} \\
        &= dh_q\Minv(q)\begin{bmatrix} \Id{n-k} \\ \Zmat{k \times
        (n-k)}\end{bmatrix} p_u + dh_q \Minv(q)\simpleB p_a
    \end{align*}
    On the constraint manifold, we have \(e = \dot{e} = 0\), which means
    \[
        dh_q\Minv(q)\simpleB p_a = dh_{p_u}\nabla_{q_u}V(q) -
            dh_q\Minv(q)\begin{bmatrix} 
            \Id{n-k} \\ \Zmat{k \times (n-k)}\end{bmatrix} p_u
    \]
    Since \(h\) is regular and \(\nabla_{q_u}\Minv(q) = 0\), we have that 
    \[
        \rank{dh_q\Minv(q)\simpleB} = k
    \]
    Solving for \(p_a\) gives
    \[
        p_a = \left(dh_q\Minv(q)\simpleB\right)\inv
        \left(dh_{p_u}\nabla_{q_u}V(q) - 
            dh_q\Minv(q)\begin{bmatrix} 
        \Id{n-k} \\ \Zmat{k \times (n-k)}\end{bmatrix} p_u \right)
    \]
    This yields a function \(p_a(q,p_u)\). However, on \(\Gamma\) we have 
    \(h(q,p_u) = 0\) and can solve for \(q_a\) in terms of \((q_u,p_u)\).
    Hence, on \(\Gamma\) we have \(p_a = g(q_u,p_u)\). Since \(q_a\) and
    \(p_a\) can be computed directly from \((q_u,p_u)\), the dynamics on
    \(\Gamma\) are parameterized only by \((\dot{q}_u,\dot{p}_u)\).
\end{proof}

Theorem \ref{thm:zero-dynamics} shows that, for a particular class of systems
and constraints, the dynamics on \(\Gamma\) are entirely described by the \(2(n-k)\)
unactuated coordinates. 
This is trueregardless of the number of degrees of freedom of the system. 
This means that \(\Gamma\) is a \(2(n-k)\)-dimensional surface inside 
\(\mathcal{Q}\times \mathcal{P}\).

The following corollay applies Theorem \ref{thm:zero-dynamics} to
systems with only one unactuated coordinate.

\begin{cor}\label{cor:2d-zero-dynamics}
    Suppose \(\mathcal{H}\) is a ---- mechanical system satisfying Assumption
    \ref{assm:H-is-simply-actuated} which has degree of underactuation one.
    Let \(h(q,p_u) = 0\) be a regular VNHC of order \((n-1)\) of the form
    \(h(q,p_u) = q_a - f(q_u,p_u)\)
    where \(f\) is a suitably-defined smooth function.
    Defining \(e_1 := (1,0,\ldots,0) \in \R^n\), the actuated momentum 
    is given by
    \begin{equation}\label{eqn:g-qupu}
        p_a = -\left.\left(dh_q \Minv(q)
        \begin{bmatrix}
            \Zmat{1\times (n-1)}\\
            I_{(n-1)}
        \end{bmatrix}\right)\inv 
        \left(\partial_{p_u}f \partial_{q_u}V + dh_q \Minv(q)e_1 p_u\right) 
            \right|_{q_a =f(q_u,p_u)}
    \end{equation}
    Since \(q_u \in \Rt{T}\) for some \(T \in ]0,\infty]\) and \(p_u \in \R\),
    the orbit \((q_u(t),p_u(t))\) traces out
    a curve on the 2D-plane \(\Rt{T} \times \R\) which we call the \((q,p)\)-plane.
\end{cor}
